scRNAseq benchmark in want of a title

### 2019-05-27 bmc_article.tex
updated manuscript...

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR). Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 14B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 15-16). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure4}A). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure4} and Supplementary Figures 15-16). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

@@

We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 17). To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically so for all dataset.

@@

We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, GLM-PCA \citep{townesGlmpca2019}, and a two-step PCA. Seurat's default PCA weighs the cell embeddings by the variance of each component,+

@@

\subsection*{Doublet detection method}
Our doublet detection method is available at \url{https://github.com/plger/plDoubletFinder}. Briefly, we first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster. We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells. We also create meta-cells from each cluster, and use them to create additional doublets and triplets. We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran. We then calculate, for each cell, the proportion of its neighbors that are artificial doublets. This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder). For the pipeline comparison here, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.

\subsection*{Filter sets}
Lenient, default, and stringent sets of filters...

>>>>

Finally, although there were differences in specific cases, we observed no robust tendency of certain normalization methods towards over- or under-estimating the number of clusters when using with Seurat clustering (Supplementary Figure 14).

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR). Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 15B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure4}A). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure4} and Supplementary Figures 16-17). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

@@

We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 18). To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically so for all dataset.

@@

We next investigated the step of dimensionality reduction, comparing in particular Seurat's PCA, GLM-PCA \citep{townesGlmpca2019}, and a two-step PCA procedure in which cells are first clustered in rough clusters according to a standard PCA, a new PCA is performed on each, and the main components of the two are put together (see methods). Seurat's default PCA weighs the cell embeddings by the variance of each component, so we also tested the impact of this weighting with each method.

The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 19). Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure 5 and Supplementary Figures 19-21). In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.

We next investigated the impact of the number of dimensions used on the clustering. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters, although the effect was milder with data processed through sctransform (Supplementary Figures 14 and 22). This tended to be translated to the accuracy of the clustering (Supplementary Figure 23), although in both cases (number of clusters and ARI) the resolution parameter had a much stronger impact.

\subsubsection*{Rules for selecting components}

\subsection*{Clustering}

@@

\subsection*{Doublet detection method}
Our doublet detection method is available at \url{https://github.com/plger/scDblFinder}. Version 0.1 was used here. Briefly, we first cluster the cells using scran's \textit{quickCluster} method, favouring overclustering (default minimum size of 20), and reduce the dataset to the union of 1500 most expressed genes of each cluster. We then create 75\% of the artifical doublets by sampling the two cells specifically from different clusters, and 25\% by entirely random sampling (in case the clustering was erroneous), and summing the counts of each pair of cells. We also create meta-cells from each cluster, and use them to create additional doublets and triplets. We then combine them with the real cells, convert the counts to scaled column ranks, and build a graph of the \textit{k} (default 10) nearest neighbors using scran. We then calculate, for each cell, the proportion of its neighbors that are artificial doublets. This ratio serves as a doublet score, which is then thresholded by simultaneously minimizing the error in classifying real vs artificial doublets and the deviation from the distribution of expected doublet rate (accounting for homotypic doublets as done by DoubletFinder). For the pipeline comparison here, the expected doublet rate was set to 2.1\% (with a 1\% standard deviation) on the basis of the mixology datasets.

\subsection*{Filter sets}
The \textit{default} set of filters excludes cells that are outliers according to at least two of the following thresholds: log10_total_counts $>$2.5 MADs or $<$5 MADs, log10_total_features $>$2.5 MADs or $<$5 MADs, pct_counts_in_top_20_features $>$ or $<$ 5 MADs, featcount_dist (distance to expected ratio of log10 counts and features) $>$ or $<$ 5 MADs, pct_counts_Mt $>$ 2.5 MADs and $>$ 0.08.

The \textit{stringent} set of filters uses the same thresholds, but excludes a cell if it is an outlier on any single distribution.

The \textit{lenient} set of filters excludes cells that are outliers on at least two distributions by at least 5 MADs, except for pct_counts_Mt where the threshold is $>$ 3 MADs and $>$ 0.08.
